---
title: "Logistic Aproximation to the Step function"
output: pdf_document
---
```{r globel_knitr_options}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=8, fig.height=6, 
               fig.alig="left",
               fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

# Introduction.
We test if two sigmoid functions can simulate a smooth curve and if Stan can estimate the change point of that curve using a logistic approximation of the step function.

## Setup
```{r library}
require(knitr)
library(tidyr)
library(rstanarm)
library(shinystan)
require(ggplot2)
library(mgcv)
library(parallel)
```

# Step Function

## Simple step function
```{r stepfunction1}
require(ggplot2)

t <- seq(from=-10, to=10, length=1000)
c <- 5

y0 <- 1/(1+ exp( -t)) 
y1 <- 1/(1+ exp( t))  

mydata<- data.frame(y0=y0, y1=y1, y2=y0 +y1, t=t)

g<- ggplot(mydata)
g +geom_point(aes(x=t, y=y0))
g +geom_point(aes(x=t, y=y1))
g +geom_point(aes(x=t, y=y2))


```

Sigmoid functions approximate the step function and cancel each other out.

## Step function with change point
```{r stepfunction2}
require(ggplot2)

t <- seq(from=-10, to=10, length=1000)

c <- 5

y0 <- 1/(1+ exp( -t+c)) 
y1 <- 1/(1+ exp(- (-t+c)))  

mydata<- data.frame(y0=y0, y1=y1, y2=y0 +y1, t=t)

g<- ggplot(mydata)
g +geom_point(aes(x=t, y=y0))
g +geom_point(aes(x=t, y=y1))
g +geom_point(aes(x=t, y=y2))


```

Stepfunctions with change point cancel each other.

# Stepfunction with slope
```{r stepfunction3}
require(ggplot2)

t <- seq(from=-10, to=0, length=1000)

c <- -5
early_slope=-1
late_slope=-5

y0 <- 1/(1+ exp(- (-t+c)))  * (early_slope *(t)) 
y1 <- (early_slope *(c)) +
  ( 1/(1+ exp( -t+c)) * late_slope *(t-c) )

y <- 1/(1+ exp(- (-t+c)))  * (early_slope *(t)) +
  (early_slope *(c)) +
  ( 1/(1+ exp( -t+c)) * late_slope *(t-c) )
  
mydata<- data.frame(y0=y0, y1=y1, y2=y0+y1, y=y, t=t)

g<- ggplot(mydata)
```

## First sigmoid
```{r}
g +geom_point(aes(x=t, y=y0))
```

## Second sigmoid
```{r}
g +geom_point(aes(x=t, y=y1))
```

## Added sigmoids
```{r}
g +geom_point(aes(x=t, y=y2)) +ylab("y1+y2")
```

## Two sigmoid functions added

```{r tow_sigmoid _function}
g +geom_point(aes(x=t, y=y)) 
```
The sum of two sigmoid functions gives a smooth curve.

# Stan model to find change point
## Generate Data
```{r generate_data}

t <- seq(from=-10, to=0, length=100)

c <- -5
early_slope=-1
late_slope=-5
sigma <-1


## to complicated
y <- 1/(1+ exp(- (-t+c)))  * (early_slope *(t)) +
  (early_slope *(c)) +
  ( 1/(1+ exp( -t+c)) * late_slope *(t-c) ) +
  rnorm(length(t),0, sigma)
  
mydata<- data.frame(y=y, t=t)
save(mydata, c, early_slope, late_slope, file="Data/mydata.Rdata")
```

## Plot data
```{r plot_data}
require(ggplot2)
load("Data/mydata.Rdata")
g<- ggplot(mydata)
g +geom_point(aes(x=t, y=y))
```

We generate data with change point c =`r c`, early slope = `r early_slope`, late slope =`r late_slope`, and error ~ N(0, 1). But these data are too complex to start with.

## Generate simple data
```{r generate__simple_data, eval=FALSE}

t <- seq(from=-10, to=0, length=100)

c <- -5

sigma <-1


## to complicated
y <- (1/(1+ exp(t-c)))
## y <- (1/(1+ exp(t)))

#y <- c*t
#y <- exp(t-c)
  
mydata2<- data.frame(y=y, t=t)
save(mydata2, c,  file="Data/mydata2.Rdata")
```

## Plot simple data
```{r plot_simple_data}
require(ggplot2)
load("Data/mydata2.Rdata")
g<- ggplot(mydata2)
g +geom_point(aes(x=t, y=y))
```

## Compile
```{r compile_z1, eval=FALSE}
require(rstan)
load("Data/mydata2.Rdata")

data_for_stan <- list(y=mydata2$y, t=mydata2$t, nObs=length(mydata2$y))
fitz1 <- stan("Stan_z1.stan", data=data_for_stan, iter=100, chains=3)
fitz2 <- stan(fit=fitz1, data=data_for_stan, iter=1000, chains=3, cores=3)
save(fitz1, fitz2, file="Fit/fitz1.Rdata")
```

## Results
```{r print_fitz1}
load("Fit/fitz1.Rdata")
print(fitz1)
print(fitz2)
```

Stan does not estimate the change point c=-5 correctly, possibly due to the non-linear transformation of the data?  We used the Stan Model in *Stan_z1.stan*. 
